Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-07-13 manuscript/main.tex
practical
reduce missing stuff

	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
	It's mathematical formulation is called the Zipfian distribution and its rank-frequency distribution is related to the inverse power law.
	Sampling $s_j$ from a Zipfian distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.
	\FIXME{TO IMPROVE, AS IT DOES NOT EXPLAIN WHY ZIPFIAN AND NOT OTHER FUNCTIONS}

@@

    \subsection{Performance}
	\FIXME{Describe the dataset (HPRC chromosomes), runtime and memory usage (specifying the fastest/slowest chromosome pangenome graph layout and the thinnest/fattest in memory), just text no with a table}
    \FIXME{A comparison with Bandage might be worth, to give readers an idea of the scalability}

    \subsection{Pangenome graphs layouts reveal biology features}
	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.
	In the supp a huge picture with a "karyotype", that is all chromosome layout together sorted by chromosome number, might be hot}

@@

	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
	\FIXME{Discuss differences to other existing methods? are there possible improvements for our method?}

>>>>

	Originating from empirical inspection of word frequency tables, Zipf's law states that a word with rank $n$ occurs $1/n$ times as the most frequent one.
	It's mathematical formulation is called the Zipfian distribution and its rank-frequency distribution is related to the inverse power law.
	Sampling $s_j$ from a Zipfian distribution fixed in the $s_i$'s path position space increases the possibility to draw a nucleotide position close to $s_i$.
	So there is a high chance to use small nucleotide distances $nd_{ij}$ to refine the layout of nodes comprising a few base pairs.

@@

    \subsection{Performance}
	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} to show the scalability and practicality of the algorithm with a real pangenome.
	Experiments were conducted on computer with \FIXME{XXX GB of RAM and XXXAMD/INTELXXX processors with XXX vCPUs}.
	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
	See \FIXME{SUPPLEMENTARY TABLE XXX WITH GRAPH LENGTH AND NUMBER OF NODES, EDGES AND STEPS} for graph statistics).
	When applied to these pangenome graphs, PG-SGD obtains the graph layouts in \FIXME{XXXX minutes/hours on average}, with the slowest chromosome being chromosome 16.
	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
	Memory consumption is \FIXME{XXXX GB of RAM on average}, with the memory peak occurring with \FIXME{chromosome XXX}.
	Given its scalability, PG-SGD can even be applied to the full graph with all chromosomes together \FIXME{SUPPLEMENTARY TABLE WITH RUNTIME, MEMORY USAGE; SUPPLEMENTARY FIGURE whe}.

    \subsection{Pangenome graphs layouts reveal biology features}
	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent.
	We show how PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure.
	\FIXME{With the HPRC dataset, Describe a full chromosome, and 1 or a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the HPRC paper or PGR-TK.}

@@

	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
